The Multi-Stencil Language

### 2016-02-01 ARTICLES/IJPP/article/parallelism.tex
article IJPP corrections+example Helene 01/02/2016

%------------------------------
\subsection{Data parallelism}
\label{sect:dataparal}
In a data parallelization technique, the idea is to split the data on which the program is computed into balanced sub-parts, one for each available resource. The same sequential program can afterwards be applied on each sub-part simultaneously, with some additioinal synchronizations between resources to update the data not computed locally, and thus to guarantee a correct result.

@@

Actually, a synchronization is needed by the data read by a stencil computation (not local), and only if this data has been modified before, which means that it has been written before. This synchronization is needed because a neighborhood function $n \in \mathcal{N}$ of a stencil computation involves values computed on different resources.

@@

When data parallelism is applied, the other type of computation which is responsible for additional synchronizations is the reduction. Actually, the reduction is applied locally on each subset of entities, on each resource, which results in $p$ (number of resources) scalar values, one on each resource. Thus, to perform the final reduction a synchronization is needed to get the $p$ scalars onto a single resource, to finally get the reduction scalar.

\begin{mydef}
For a reduction kernel $k_j(S_j,R_j,(w_j,D_j),comp)$, if we denote by $w^r$, $0 \leq r<p$, the local $w$ computed on the resource $r$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
k_{j}^{sync}(S,R,(w,D),comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$, and $w=w_i$, $D=entity(w)=D_i$ and $comp=comp_i$.

@@

\begin{algorithm}
\caption{$F_{sync}$ recursive function}
\label{alg:sync}
\begin{algorithmic}[1]
\Procedure{$F_{sync}$} {$\Gamma$,$j$}
\State $k_j = \Gamma[j]$
\State $list = []$
\If {$j=|\Gamma|$}
\State return $list$
\ElsIf {$\exists (r_j,n_j) \in R_j$ such that $n_j\neq identity$}
\For {$0 \leq i<j$}
\State $k_i = \Gamma[i]$
\If {$k_i \pprec k_j$}
\State $S = \emptyset$
\State $R = \{(w_i,\bigcup_{e \in D_i} n_j(e))\}$
\State $(w,D) = (w_i,\bigcup_{e \in D_i} n_j(e)))$
\State $comp = identity$
\State $list.[k_{i;j}^{sync}(S,R,(w,D),comp)$
\State break
\EndIf
\EndFor
\State $list \cdot [k_j]$
\ElsIf {$w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}$}
\State $S = \emptyset$
\State $R = \{(w^0_j,entity(w^0_j)) \dots (w^{p-1}_j,entity(w^{p-1}_j))\}$
\State $(w,D) = (w_j,entity(w_j))$
\State $comp = comp_j$
\State $list.[k_j, k_{j}^{sync}(S,R,(w,D),comp)]$
\Else
\State $list.[k_j]$
\EndIf
\State return $list \cdot F_{sync}(\Gamma,j+1)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq m \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\mathcal{N},\Delta_m,\mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.

%------------------------------
\subsection{Hybrid parallelism}
A task parallelization technique is a technique to transform a program as a dependency graph of different tasks. A dependency graph exhibits parallel tasks, or on the contrary sequential execution of tasks. Such a dependency graph can directly be given to a dynamic scheduler, or can be statically scheduled. In this paper, we introduce task parallelism by building the dependency graph between computations of the sequential list $\Gamma_{sync}$ (which itself take into account data parallelism).

@@

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computation kernels, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $F_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function

\begin{algorithm}
\caption{$F_{dep}$ recursive function}
\label{alg:sync}
\begin{algorithmic}[1]
\Procedure{$F_{dep}$} {$\Gamma_{sync}$,$j$}
\State $k_j = \Gamma_{sync}[j]$
\If {$j=|\Gamma_{sync}|$}
\State return $(\{\},\{\})$
\ElsIf {$j<|\Gamma_{sync}|$}
\State $G=(\{\},\{\})$
\For {$0 \leq i<j$}
\State $k_i = \Gamma_{sync}[i]$
\If {$k_i \prec_{rw} k_j$ or $k_i \prec_{w} k_j$}
\State $G = G \cup (k_j, \{(\overset{\frown}{k_i,k_j} \})$
\EndIf
\EndFor
\State return $G \cup F_{dep}(\Gamma_{sync},j+1)$
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{proof}
$\Gamma_{dep}$ is built from $\Gamma_{sync}$ which is an ordered and sequential list of computations. Moreover, each computation of the list $\Gamma_{sync}$ is associated to a vertex of $V$, even if the same computation is represented more than once in $\Gamma_{sync}$. As a result it is not possible to go back to a previous computation and to create a cycle.
\end{proof}

As a result of the hybrid parallelization, each resource $0 \leq m \leq p-1$ perform a multi-stencil program, defined by
\begin{equation*}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\mathcal{N},\Delta_m,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and synchronizations of kernels added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}

>>>>

%------------------------------
\subsection{Data parallelism}
\label{sect:dataparal}
In a data parallelization technique, the idea is to split quantities on which the program is computed into balanced sub-parts, one for each available resource. The same sequential program can afterwards be applied on each sub-part simultaneously, with some additioinal synchronizations between resources to update the data not computed locally, and thus to guarantee a correct result.

@@

Actually, a synchronization is needed by the data read by a stencil computation (not local), if this data has been modified before, which means that it has been written before. This synchronization is needed because a neighborhood function $n \in \mathcal{N}$ of a stencil computation involves values computed on different resources.

@@

When data parallelism is applied, the other type of computation which is responsible for additional synchronizations is the reduction. Actually, the reduction is first applied locally on each subset of entities, on each resource. This results in $p$ (number of resources) scalar values. Thus, to perform the final reduction a synchronization is needed to get the $p$ scalars onto each resource. The final reduction step is applied such that each resource get the result scalar.

\begin{mydef}
For a reduction kernel $k_j(S_j,R_j,(w_j,D_j),comp)$, if we denote by $w^r$, $0 \leq r<p$, the local scalar $w$ computed on the resource $r$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
k_{j}^{sync}(S,R,(w,D),comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$, and $w=w_i$, $D=entity(w)=D_i$ and $comp=comp_i$.

@@

@@

\begin{algorithm}
\caption{$F_{sync}$ recursive function}
\label{alg:sync}
\begin{algorithmic}[1]
\Procedure{$F_{sync}$} {$\Gamma$,$j$}
\State $k_j = \Gamma[j]$
\State $list = []$
\If {$j=|\Gamma|$}
\State return $list$
\ElsIf {$\exists (r_j,n_j) \in R_j$ such that $n_j\neq identity$}
\For {$0 \leq i<j$}
\State $k_i = \Gamma[i]$
\If {$k_i \pprec k_j$}
\State $S = \emptyset$
\State $R = \{(w_i,\bigcup_{e \in D_i} n_j(e))\}$
\State $(w,D) = (w_i,\bigcup_{e \in D_i} n_j(e)))$
\State $comp = identity$
\State $list.[k_{i;j}^{sync}(S,R,(w,D),comp)$
\EndIf
\EndFor
\State $list \cdot [k_j]$
\ElsIf {$w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}$}
\State $S = \emptyset$
\State $R = \{(w^0_j,entity(w^0_j)) \dots (w^{p-1}_j,entity(w^{p-1}_j))\}$
\State $(w,D) = (w_j,entity(w_j))$
\State $comp = comp_j$
\State $list.[k_j, k_{j}^{sync}(S,R,(w,D),comp)]$
\Else
\State $list.[k_j]$
\EndIf
\State return $list \cdot F_{sync}(\Gamma,j+1)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq r \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_r(T,\mathcal{M}_r,\mathcal{E}_r,\mathcal{D}_r,\mathcal{N},\Delta_r,\mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.

%------------------------------
\subsection{Hybrid parallelism}
A task parallelization technique is a technique to transform a program as a dependency graph of different tasks. A dependency graph exhibits parallel tasks, or on the contrary sequential execution of tasks. Such a dependency graph can directly be given to a dynamic scheduler, or can statically be scheduled. In this paper, we introduce task parallelism by building the dependency graph between kernels of the sequential list $\Gamma_{sync}$. Thus, as $\Gamma_{sync}$ takes into account data parallelism, we introduce hybrid parallelism.

@@

Those definitions are close to traditionnal read/write dependencies. However, specific conditions due to the specific domain can be introduced, such as the condition on boundary kernels, and conditions on the computation domain.

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computation kernels, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $F_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function defined in Algorithm~\ref{alg:dep}.

@@

\begin{algorithm}
\caption{$F_{dep}$ recursive function}
\label{alg:dep}
\begin{algorithmic}[1]
\Procedure{$F_{dep}$} {$\Gamma_{sync}$,$j$}
\State $k_j = \Gamma_{sync}[j]$
\If {$j=|\Gamma_{sync}|$}
\State return $(\{\},\{\})$
\ElsIf {$j<|\Gamma_{sync}|$}
\State $G=(\{\},\{\})$
\For {$0 \leq i<j$}
\State $k_i = \Gamma_{sync}[i]$
\If {$k_i \prec_{rw} k_j$ or $k_i \prec_{w} k_j$}
\State $G = G \cup (k_j, \{(\overset{\frown}{k_i,k_j} \})$
\EndIf
\EndFor
\State return $G \cup F_{dep}(\Gamma_{sync},j+1)$
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

% \begin{proof}
% $\Gamma_{dep}$ is built from $\Gamma_{sync}$ which is an ordered and sequential list of computations. Moreover, each computation of the list $\Gamma_{sync}$ is associated to a vertex of $V$, even if the same computation is represented more than once in $\Gamma_{sync}$. As a result it is not possible to go back to a previous computation and to create a cycle.
% \end{proof}

As a result of the hybrid parallelization, each resource $0 \leq r \leq p-1$ perform a multi-stencil program, defined by
\begin{equation*}
\mathcal{MSP}_r(T,\mathcal{M}_r,\mathcal{E}_r,\mathcal{D}_r,\mathcal{N},\Delta_r,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computation kernels $k_i$ of $\Gamma$ and synchronizations of kernels added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}
